\section{Renewal Reward Theorem and load}
\label{sec:renew-reward-theor}

\Opensolutionfile{hint}
\Opensolutionfile{ans}
\subsection*{Theory and Exercises}

We start with stating and proving (graphically) the \recall{renewal reward theorem}.
In the sequel we will see many applications of this theorem.
The first application is to relate the fraction of time the server is busy in a $G/G/1$ queue to the job arrival rate and the expected job service time.


The renewal reward theorem is very useful, and states intuitively that when customers arrive at rate $\lambda$ and each customer pays an average amount $X$, then the system earns money at rate $Y=\lambda X$.
Figure~\ref{fig:renewal} provides graphical motivation about why this theorem is true; \citet{el-taha98:_sampl_path_analy_queuein_system} gives a (simple) proof.

\begin{theorem}[Renewal Reward Theorem, $Y=\lambda X$]
  Consider epochs $\{T_k, k=0, 1, \ldots\}$ such that  $0=T_0 < T_1 < \cdots$
  Let $N=\{N(t), t\geq 0\}$ be the associated counting process with $N(t) = \max\{k : T_k \leq t\}$.
  Let $\{Y(t), t\geq 0\}$ be a non-decreasing right-continuous (deterministic) process.
  Define $X_k = Y(T_k)-Y(T_{k-1})$.
  Suppose that $N(t)/t\to\lambda$ as $t\to\infty$, where $0<\lambda < \infty$.
  Then $Y(t)/t$ has a limit iff $n^{-1}\sum_k^n X_k$ has a limit, and then $Y=\lambda X$. In other words, 
 \begin{equation*}
   \lim_{t \to \infty} \frac{Y(t)}t=Y \iff \lim_{n \to \infty} \frac 1n\sum_{k=1}^n X_k = X, 
 \end{equation*}
and then $Y=\lambda X$. 
\end{theorem}


\begin{figure}[ht]
  \centering
\begin{tikzpicture}[scale=1]
%axis
\draw[->] (0,0) -- coordinate (x axis mid) (8.5,0);
\draw[->] (0,0) -- coordinate (y axis mid) (0,5.5);
%\node[below=0.2cm] at (x axis mid) {$t$};

\draw plot [smooth] coordinates {(1,0.7) (2,1) (3,1.8) (4,2) (5,3) (6,3.2) (7, 3.4) (7.5,4.0)};
\node[right]  at (7.1,3.4) {$Y(t)$};
\node  at (7.,-0.5) {$t$};

\node  at (2.,-0.5) {$T_{k-1}$};
\draw[dotted] (2,0)--(2,1);
\draw[dotted] (2,1)--(3,1);


\node  at (3.,-0.5) {$T_{k}$};
\draw[dotted] (3,0)--(3,1.8);
\draw[dotted] (3,1.8)--(5,1.8);

\draw [
    thick,
    decoration={brace, mirror, raise=0.1cm },
    decorate
] (3,1) -- (3,1.8)
node[pos=0.5, xshift=0.6cm]  {$X_k$}; 


\node  at (5.,-0.5) {$T_{k+1}$};
\draw[dotted] (5,0)--(5,3);

\draw [
    thick,
    decoration={brace, mirror, raise=0.1cm },
    decorate
] (5,1.8) -- (5,3)
node[pos=0.5, xshift=0.7cm]  {$X_{k+1}$}; 

\node[right] at (0.5,4) {$\frac{Y(t)}t \approx \frac{\sum_i^n X_i}{\sum_{i}^n T_i} = \frac{n}{\sum_{i}^n T_i} \frac{\sum_i^n X_i}n $};


\draw[dotted,<->, =stealth] (7,0.05)--(7,3.35) node[midway, rotate=90, fill=white] {$\sum_{i=1}^n X_i$};
\draw[dotted,<->, =stealth] (0.05,0.5)--(6.95,0.5) node[pos=0.85, fill=white] {$\sum_{i=1}^n T_i$};
\end{tikzpicture}
\caption{A graphical `proof' of $Y=\lambda X$. Here $Y(t)/t\to Y$, $n/\sum^n_i T_i\to \lambda$ and $n^{-1}\sum_i^n X_i \to X$.  (Observe that in the
  figure $X_k$ does not represent an inter-arrival time; instead it
  corresponds to the increment of (the graph of) $Y(t)$ between two
  consecutive epochs $T_{k-1}$ and $T_k$ at which $Y(t)$ is
  observed.) }
  \label{fig:renewal}
\end{figure}

Define the \recall{load} or \recall{utilization} as the limiting fraction of time the server is busy, i.e.,
\begin{equation*}
  \rho = \lim_{t\to\infty} \frac 1 t \int_0^t \1{L(s)>0} \d s.
\end{equation*}

\begin{exercise}[\faPhoto]
Use the renewal reward theorem to prove that $\rho = \lambda \E S$.
\begin{hint}
  Define $Y(t) = \int_0^t \1{L(s)>0} \d s$ as the total amount of time the server has been busy up to the time $t$.
  Then take as epochs $T_k = D_k$ and use rate stability.
\end{hint}
\begin{solution}
  It is evident that $X_k =  Y(D_k)-Y(D_{k-1})=S_k$, hence $X = \lim_{n\to\infty} n^{-1}\sum_{k=1}^n X_k = \E S$.
  Also $\lim_{t\to\infty} Y(t)/t=\rho$.
  Finally, in the relation $Y = \lambda X$, the $\lambda$ is $\delta$ since we consider departure epochs $T_k = D_k$, rather than $A_k$.
  By the renewal reward theorem $Y=\lambda X$ we get that $\rho = \delta \E S$.
  Finally, by rate-stability, the job arrival rate $\lambda = \delta$, hence $\rho = \lambda \E S$.
\end{solution}
\end{exercise}


\begin{exercise}[\faCalculator]
  We can derive the relation $\rho = \lambda \E S$ in a somewhat more direct way by considering the fact that
\begin{equation*}
  \sum_{k=1}^{A(t)} S_k \geq \int_0^t \1{L(s)>0} \d s \geq   \sum_{k=1}^{D(t)} S_k.
\end{equation*}
Explain this, and complete the argument.
\begin{solution}
Observe that
since $t$ can lie half way a service interval and $A(t) \geq D(t)$. As
$A(t)\to \infty$ as $t\to\infty$,
\begin{equation*}
  \lim_{t\to\infty} \frac 1 t\sum_{k=1}^{A(t)} S_k = 
  \lim_{t\to\infty} \frac{A(t)}t \frac{1}{A(t)} \sum_{k=1}^{A(t)} S_k = 
  \lim_{t\to\infty} \frac{A(t)}t \cdot \lim_{t\to\infty}\frac{1}{A(t)} \sum_{k=1}^{A(t)} S_k = \lambda \E S.
\end{equation*}
Applying similar limits to the other inequality gives
\begin{equation*}
\lambda \E S \geq \rho \geq \delta \E S.
\end{equation*}
Hence, if $\delta=\lambda$, $\rho = \lambda \E S$.  

Note that this is in fact the same argument as that underlies the renewal reward theorem. Henceforth we will just use the renewal reward theorem. 
\end{solution}
\end{exercise}


From the identities $\lambda^{-1} = \E X$ and $\mu^{-1} = \E S$, we get
a further set of relations:
\begin{equation*}
  \rho = \lambda \E S = \frac{\lambda}{\mu} = \frac{\E S}{\E X}.
\end{equation*}
Thus, the load has also the interpretation as the rate at which jobs
arrive times the average amount of work per job.  Finally,
recall that for a system to be rate-stable, it is necessary that
$\mu> \lambda$, implying in turn that $\rho < 1$. The relation
$\rho=\E S/ \E X < 1$ then tells us that the average time it takes to
serve a job must be less than the average time between two consecutive
arrivals, i.e., $\E S < \E X$. In fact, when $\mu < \lambda$, it is easy to check with simulation that the queue length
grows roughly linearly with slope $\lambda - \mu$.


\begin{exercise}[\faFlask]
  Consider a queueing system with $c$ identical servers (identical in
  the sense that each server has the same production rate $\mu$). What would be a reasonable stability criterion for this system? 
  \begin{hint}
What is the rate in, and what is the service capacity?
  \end{hint}
  \begin{solution}
    The criterion is that $c$ must be such that $\lambda <
    c\mu$.
    (Thus, we interpret the number of servers as a \emph{control},
    i.e., a `thing' we can change, while we assume that $\lambda$ and
    $\mu$ cannot be easily changed.) To see this, we can take two
  different points of view.  Imagine that the $c$ servers are replaced
  by one server that works $c$ times as fast. The service capacity of
  these two systems (i.e., the system with $c$ servers and the system
  with one fast server) is the same, i.e., $c\mu$, where $\mu$ is the
  rate of one server. For the system with the fast server the load is
  defined as $\rho =\lambda/c\mu$, and for stability we require
  $\rho<1$.  Another way to see it is to assume that the stream of
  jobs is split into $c$ smaller streams, each with arrival rate
  $\lambda/c$.  In this case, applying the condition that
  $(\lambda/c )/\mu<1$ per server leads to the same condition that
  $\lambda/(c\mu) < 1$.
  \end{solution}
\end{exercise}

\begin{exercise}[\faPhoto]\label{ex:11}
  Define a \recall{busy time} as the moment a job arrives at an empty system and the server starts serving until the server becomes idle again.
  Let $\E{B}$ be the expected busy time and $\E{I}$  the expected idle time. Show that $\E{B} = \E{I} \rho/(1-\rho) $ is the fraction of time the server is busy. Interpret the result.
  \begin{hint}
    Use the concept of busy cycle developed in Exercise~\ref{ex:29}. 
  \end{hint}
\begin{solution}
  Consider again a busy cycle, that is, a cycle that starts with the first
  job that sees an empty system at upon arrival up to the time another
  job sees an empty system upon arrival. In such one cycle, the server
  is busy for an expected duration $\E{B}$. The total expected length
  of the cycle is $\E{B} + \E{I}$, since after the last job of the cycle
  left, the expected time until the next job is $\E{I}$.

  Since $\rho$ is the utilization of the server,
  \begin{equation*}
\rho = \frac{\E B}{\E B + \E I}.
  \end{equation*}
  With a bit of algebra the result follows.

  The interpretation is that once we know $\rho$, the ratio between $\E B$ and $\E I$ is fixed. 
\end{solution}
\end{exercise}

\Closesolutionfile{hint}
\Closesolutionfile{ans}

\opt{solutionfiles}{
\subsection*{Hints}
\input{hint}
\subsection*{Solutions}
\input{ans}
}
%\clearpage


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../queueing_book"
%%% End:
